Intrinsic Verification of Parsers and Formal Grammar Theory in Dependent Lambek Calculus

### 2024-11-15 paper/paper.tex
add a flag for the appendix

@@

\pagestyle{plain}

@@

Finite automata are precisely the class of machines that recognize regular
expressions. \cref{fig:exampleNFA} shows a non-deterministic
finite automaton (NFA) for the regular expression
$({\literal a}^{*} \otimes \literal b) \oplus \literal c$, along with a type $\Trace$, an \emph{indexed} inductive linear type of traces through this automaton. Defining an indexed inductive type can be thought
of as defining a family of mutually recursive inductive types, one for each element of the indexing type. Here $\Trace$ 
uses an index $\s : \texttt{Fin 3}$ which picks out which
state in the automaton a trace begins at --- where $\texttt{Fin 3}$ is the
finite type containing inhabitants $\{\0 , \1 , \2\}$. We can think of this as defining three mutually recursive inductive types $\Trace~\0$,
$\Trace~\1$, and $\Trace~\2$.
%
There are three kinds of constructors for $\Trace$: (1) those that
terminate traces, (2) those that correspond to transitions labeled by
a character, and (3) those that correspond to transitions labeled by
the empty string $\epsilon$. The constructor $\texttt{stop}$
terminates a trace in the accepting state $\2$. The constructors
$\texttt{1to1}$, $\texttt{1to2}$, $\texttt{0to2}$ each define a
labeled transition through the NFA, and each of these consumes a parse
of the label's character and a trace beginning at the destination of a
transition to produce a trace beginning at the source of a
transition. The constructor $\texttt{0to1}$ behaves similarly, except
its transition is labeled with the empty string $\epsilon$. Therefore,
$\texttt{0to1}$ takes in a trace beginning a state $\1$ and returns a
trace beginning at state $\0$ corresponding to the same underlying
string.
%
Lastly, we give a lambda term that constructs an accepting trace
starting at the initial state for the string \stringquote{ab}.  Later
in \Cref{sec:applications}, we will show that we can actually
construct mutually inverse functions between the regular expression
$({\literal a}^{*} \otimes \literal b) \oplus \literal c$ and its
corresponding NFA traces ($\Trace~\0$) demonstrating that the regular
expression and the automaton capture the same language. Further since
the functions are mutually inverse, this shows they are \emph{strongly
equivalent} as grammars.

@@

The design of \theoryabbv is based the dependent
linear-non-linear calculus (\lnld) and Lambek calculus, also known as
non-commutative linear logic.
\cite{krishnaswami_integrating_2015,lambek58}. As in \lnld,
\theoryabbv includes both non-linear dependent types, as well as
linear types, which are allowed to depend on the non-linear types, but
not on other linear types.
%
The main point of departure from \lnld's design is that, as in Lambek calculus \cite{lambek58}, the linear
typing is \emph{non-commutative}, i.e., that exchange is not an
admissible structural rule. Furthermore, we add a general-purpose
indexed inductive linear type connective, as well as an
\emph{equalizer} type, which we will show allows us to perform
inductive proofs of equalities between linear terms.
%
Finally, while \lnld was enhanced with special connectives
inspired by separation logic to model imperative programming, we
instead add base types and axioms to the system specifically to model
formal grammars and parsing.

The formation rules for the judgments of \theoryabbv are shown in
\Cref{fig:formation}. $\Gamma$ stand for non-linear contexts, $X,Y,Z$
stand for non-linear types, $M,N$ stand for non-linear terms, these
act as in an ordinary dependent type theory. $\Delta$ stands for
linear contexts, $A,B,C$ for linear types and $e,f,g$ for linear
terms. These contexts, types and terms are allowed to depend on an
ambient non-linear context $\Gamma$, but note that linear types $A$
cannot depend on any \emph{linear} variables in $\Delta$. We include
definitional equality judgments for both kinds of type and term
judgments as well. Additionally, we have a judgment $\Gamma\vdash X
\isSmall$ which is used in the definition of universe types.

@@

    \inferrule{\Gamma; \Delta \vdash e : A \and \Gamma;\Delta \vdash f : A}{\Gamma;\Delta\vdash e \equiv f : A}
  \end{mathpar}
  \caption{Formation Rules}
  \label{fig:formation}
\end{figure}

@@

In addition to these type-theoretic principles, we need two additional
axioms that do not generally hold in systems based on linear
logic. First, we need that additive conjunction \emph{distributes}
over additive disjunction, e.g., in the finitary case that $0 \& A
\cong 0$ and $(A + B) \& C \cong (A \& C) + (B \& C)$. In its most
general form, the axiom says that the definable function
${\bigoplus\limits_{f : \prod_{x:X}
    Y(x)}\bigamp\limits_{x:X}A\,x\,(f\,x) \lto
  \bigamp\limits_{x:X}\bigoplus\limits_{y:Y(x)}A\,x\,y}$ has an
inverse. Second, we need that the different constructors of
$\bigoplus$ are disjoint. We add this by adding the axiom that for any
$A : X \to L$ and $x \neq x' : X$ there is a function
$\uparrow(\equalizer{b}{\sigma\,x \circ \pi_1}{\sigma\,x'\circ \pi_2}
\lto 0)$ where $b: A(x) \& A(x')$, i.e., the grammar of pairs of an
$a:A(x)$ and an $a': A(x')$ such that $\sigma\,x\,a = \sigma\,x'\,a'$
is empty.

@@

\begin{figure}
  \begin{footnotesize}
  \begin{mathpar}
    \inferrule{}{\Gamma \vdash \I \isLinTy}\and
    \inferrule{\Gamma \vdash A \isLinTy \and \Gamma \vdash B \isLinTy}{\Gamma \vdash A \otimes B \isLinTy}\and
    \inferrule{\Gamma \vdash A \isLinTy \and \Gamma \vdash B \isLinTy}{\Gamma \vdash A \lto B \isLinTy}\and
    \inferrule{\Gamma \vdash A \isLinTy \and \Gamma \vdash B \isLinTy}{\Gamma \vdash A \tol B \isLinTy}\and
    \inferrule{\Gamma,x:X \vdash A \isLinTy}{\Gamma \vdash \bigoplus\limits_{x:X} A \isLinTy}\and
    \inferrule{\Gamma,x:X \vdash A \isLinTy}{\Gamma \vdash \bigamp\limits_{x:X} A \isLinTy}\and
    \inferrule
    {\Gamma \vdash f : \ltonl {(A \lto B)} \and \Gamma \vdash g : \ltonl { (A \lto B) }}
    {\Gamma \vdash \equalizer {a}{f}{g} \isLinTy}
  \end{mathpar}
\begin{mathpar}
    \inferrule{~}{\Gamma ; a : A \vdash a : A}
    \and
    \inferrule{\Gamma ; \Delta \vdash e : B \\ \linctxwffjdg \Gamma {A \equiv B}}{\Gamma ; \Delta \vdash e : A}
    %
    \\
    %
    \inferrule{~}{\Gamma ; \cdot \vdash () : I}
    \and
    \inferrule{\Gamma ; \Delta_2 \vdash e : I \\ \Gamma ; \Delta_1,\Delta_3 \vdash e' : C}{\Gamma ; \Delta_1,\Delta_2,\Delta_3 \vdash \letin {()} e {e'} : C}
    %
    \\
    %
    \inferrule{\Gamma ; \Delta \vdash e : A \\ \Gamma ; \Delta' \vdash e' : B}{\Gamma ; \Delta, \Delta' \vdash (e , e') : A \otimes B}
    %
    \and
    %
    \inferrule{\Gamma ; \Delta_2 \vdash e : A \otimes B \\ \Gamma ; \Delta_1, a : A, b : B, \Delta_2 \vdash e' : C}{\Gamma ;  \Delta_1, \Delta_2, \Delta_3 \vdash \letin {(a , b)} e {e'} : C}
    \\
    %
    \inferrule{\Gamma ; \Delta , a : A \vdash e : B}{\Gamma ; \Delta \vdash \lamblto a e : A\lto B}
    \and
    \inferrule{\Gamma ; \Delta' \vdash e' : A \\ \Gamma ; \Delta \vdash e : A \lto B}{\Gamma ; \Delta, \Delta' \vdash \applto {e'} {e} : B}
    \\
    %
    \inferrule{\Gamma, x : X ; \Delta  \vdash e : A}
              {\Gamma ; \Delta \vdash \dlamb x e : \LinPiTy x X A}
    \and
    \inferrule{\Gamma ; \Delta \vdash e : \LinPiTy x X A \\ \Gamma \vdash M : X}{\Gamma ; \Delta \vdash e\,.\pi\,M : \subst A {M} x}
    %
    \\
    %
    \inferrule{\Gamma \vdash M : X \quad \Gamma ; \Delta \vdash e : \subst A M x}{\Gamma ; \Delta \vdash \sigma\,M\,e : \bigoplus\limits_{x:X} A}
    %
    \and
    %
    \inferrule{\Gamma ; \Delta_2 \vdash e : \bigoplus\limits_{x:X} A \quad \Gamma, x : X ; \Delta_1, a : A, \Delta_3 \vdash e' : C}{\Gamma; \Delta_1, \Delta_2, \Delta_3 \vdash \letin {\sigma\,x\,a} e {e'}: C}
    %
    \and
    %
    \inferrule
    { \Gamma ; \Delta \vdash e : A \\
      \Gamma ; \Delta \vdash \applto {f}{e} \equiv \applto {g}{e}}
    {\Gamma ; \Delta \vdash \equalizerin{e} : \equalizer {a}{f}{g}}
    %
    \and
    %
    \inferrule
    {\Gamma ; \Delta \vdash e : \equalizer{a}{f}{g}}
    {\Gamma ; \Delta \vdash \equalizerpi {e} : A}
  \end{mathpar}    
  \end{footnotesize}
  \caption{Linear Types and Terms (selection)}
  \label{fig:linear-types-and-terms}
\end{figure}

@@

    \inferrule*[right=Ind$\eta$]
    {\Gamma \vdash f : \prod_{x:X}\ltonl{(\el(F\,x)(A) \lto A\,x)}
     \and \Gamma \vdash e : \prod_{x:X}\ltonl{(\mu F\,x \lto A\,x)}\\\\
      \Gamma,x:X;a:\el(F\,x)(\mu F) \vdash e\,x\,(\roll a) \equiv f\,x\,(\map(F\,x)\,e) : A\,x }
    {\fold\,F\,f\equiv e' : \prod_{x:X}\ltonl{(\mu F\,x \lto A\,x)}}
  \end{mathpar}
  \end{footnotesize}
  %% \begin{footnotesize}
  %%   \begin{align*}
  %%   \mathsf{Var} &: \prod_{X:U} X \to \SPF\,X\\
  %%   \mathsf{K} &: \prod_{X:U} L \to \SPF\,X\\
  %%   \mathsf{\bigoplus} &: \prod_{X:U}\prod_{Y:U}(Y \to \SPF\,X) \to \SPF\,X\\
  %%   \mathsf{\bigamp} &: \prod_{X:U}\prod_{Y:U}(Y \to \SPF\,X) \to \SPF\,X\\
  %%   \mathsf{\otimes} &: \prod_{X:U}\SPF\,X \to \SPF\,X \to \SPF\,X\\
  %%   \el &: \prod_{X:U}\SPF\,X \to (X \to A) \to L\\
  %%   \map &: \prod_{X:U}\prod_{F : \SPF\,X}\prod_{A,B:X\to L}{\left(\prod_{x:X}\uparrow(\unquoteTy{A\,x}\lto \unquoteTy{B\,x})\right)} \to {\uparrow(\unquoteTy{\el(F)(A)} \lto \unquoteTy{\el(F)(B)})}\\\\
  %%   \roll &: \prod_{X:U}\prod_{F:X \to \SPF\,X}\prod_{x:X}\ltonl{(\el(F\,x)(\mu\,F))}\\
  %%   \fold &: \prod_{X:U}\prod_{F:X\to\SPF\,X}\prod_{A:X \to L}
  %%   \left(\prod_{x:X}\ltonl{(\el(F\,x)(A) \lto A\,x)}\right)
  %%   \to \prod_{x:X}\ltonl{(\mu F\,x \lto A\,x)}
  %% \end{align*}
  %% \end{footnotesize}
  \caption{Strictly Positive Functors and Indexed Inductive Linear Types}
  \label{fig:iilt}
\end{figure}

@@

%% \paragraph{Weak and Strong Equivalence}
In the theory of formal grammars, there are two different notions of
equivalence: weak generative capacity, meaning just which strings are
accepted by the grammar, and \emph{strong} generative capacity, when
the parse trees of the two grammars are isomorphic
\cite{chom1963}. Using linear types as grammars, we can define both of
these notions of equivalence in \theoryabbv.
\begin{definition}
  \label{def:weakequiv}
  Grammars $\A$ and $\B$ are \emph{weakly equivalent} if there exists
  parse transformers $\f : \ltonl{(\A \lto \B)}$ and $\g : \ltonl {(\B
    \lto \A)}$. $\A$ is a \emph{retract} of $\B$ if $\A$ and $\B$ are
  weakly equivalent and $\lambda a. g(f(a)) \equiv \lambda a.a$. They
  are \emph{strongly equivalent} if further the other composition is
  the identity, i.e., $\lambda b. f(g(b)) \equiv \lambda b.b$.
\end{definition}

@@

%% \paragraph{Ambiguity}
A formal grammar $\A$ is ambiguous if there are multiple parse trees
for the \emph{same} string $\w$.  For example, $\a \oplus \a$ is
ambiguous because there are two parses of $\stringquote{a}$,
constructed using $\inl$ and $\inr$. On the other hand, a formal
grammar is unambiguous when there is at most one parse tree for any
input string. We can capture this notion as a type in \theoryabbv in a
clever way:
\begin{definition}
  \label{def:unambig}
  A grammar $\A$ is \emph{unambiguous} if for every linear type $\B$,
  $\f : \ltonl{(\B \lto \A)}$, and $\g : \ltonl{(\B \lto \A)}$
  then $\f \equiv \g$
\end{definition}
\cref{def:unambig} can be read more intuitively as stating that $\A$
is unambiguous if there is at most one way to transform parses of any
other grammar $\B$ into parse of $\A$. This notion of an unambiguous
type is the analog for linear types of the definition of a (homotopy)
\emph{proposition} in the terminology of homotopy type
theory\cite{hottbook}. The most basic unambiguous types are $\top$ and
$0$, and in a system of classical logic all unambiguous types would
have to be equivalent to one of these, but with our axioms we can show
also that $\I$ and literals $\literal c$ are unambiguous. To see this, first, we
establish two useful properties of unambiguity.
\begin{lemma}[\agdalogo]
  If $\B$ is unambiguous and $\A$ is a retract of $\B$ then $\A$ is unambiguous.
  If a disjunction $\bigoplus\limits_{x:X} A(x)$ is unambiguous then
  each $A(x)$ is unambiguous.
\end{lemma}
From the first principle, we can prove that $\String$ is unambiguous,
since it is a retract of $\top$. In fact, observe that if $\A$ is a
retract of $\B$ and $\B$ is unambiguous, then in fact $\A$ and $\B$
are strongly equivalent, as the equation $\lambda b. f(g(b)) \equiv
\lambda b. b$ follows because $\B$ is unambiguous. Therefore
$\String$ is also strongly equivalent to $\top$. Next, since $\String$ is
defined as a Kleene star, we can easily show that $\String \cong \I
\oplus \Char \oplus (\Char \otimes \Char \otimes \String)$. Then by
the second principle we have that $\I$ and $\Char$ and each literal
$\literal c$ are unambiguous as well.

@@

We now turn to our main task, which is using our linear type system to
implement verified parsers. Given a grammar defined as a linear type
$A$, a first attempt at defining a parser would be to implement a
function $\uparrow{(\String \lto A)}$. But since our linear functions
must be total, this means that we can construct an $A$ parse for
\emph{every} input string, which is impossible for most grammars of
interest. Instead we might try to write a partial function as a
$\ltonl{(\String \lto (A \oplus \top))}$ using the ``option''
monad. This allows for the possibility that the input string doesn't
parse, but is far too weak as a specification: we can trivially
implement a parser for any type by always returning $\inr$. The
correct notion of a parser should be one that allows for failure, but
only in the case that a parse cannot be constructed.
\begin{definition}
  A parser for a linear type $A$ is a function $\uparrow{(\String \lto
    A \oplus A_{\neg})}$ where $A_{\neg}$ is a linear type that is
  \emph{disjoint} from $A$ in that we can implement a function
  $\uparrow{(A \& A_{\neg} \lto 0)}$
\end{definition}
Here we replace $\top$ in our partial parser type with a type
$A_{\neg}$ that we can think of as a negation of $A$. The function
$\uparrow{(A \& A_{\neg} \lto 0)}$ ensures that it is impossible for
$A$ and $A_{\neg}$ to match the same input string. This means that in
defining a parser, we will need to define a kind of negative grammar
for strings that do not parse. Fortunately, we will see that
deterministic automata naturally support such a notion with no
additional effort: the negative grammar is simply the grammar for
traces that end in a rejecting state. This follows from the following
principle, a consequence of our axiom that $\oplus$ constructors are
disjoint.
\begin{lemma}[\agdalogo]
  If $A \oplus A_{\neg}$ is unambiguous, then $A$ and $A_{\neg}$ are
  disjoint.
\end{lemma}

@@

print(*@$_D$@*) : (s : D.states) (*@$\to$@*) (*@$\uparrow$@*)(((*@$\oplus$@*)[ b : Bool ] Trace(*@$_{\D}$@*) s b) (*@$\lto$@*) String)
print(*@$_D$@*) s ((*@$\sigma$@*) b (Trace(*@$_{\D}$@*).nil .s)) = String.nil
print(*@$_D$@*) s ((*@$\sigma$@*) b (Trace(*@$_{\D}$@*).cons c (D.(*@$\delta$@*) c .s) b a trace)) =
  String.cons ((*@$\sigma$@*) c a) (print(*@$_D$@*) (D.(*@$\delta$@*) c s) ((*@$\sigma$@*) b trace))
\end{lstlisting}
\caption{Parser/Printer for DFA traces}
\label{fig:printdfa}
\end{figure}

@@

To justify our assertion that \theoryabbv is a syntax for formal
grammars and parse transformers, we will now define a
\emph{denotational semantics} that makes this mathematically precise
by defining a notion of formal grammar and parse transformer and
showing that our type theory can be soundly interpreted in this
model. We then discuss how this denotational semantics provides the
basis for our prototype implementation in Agda.

>>>>

\newif\ifappendix
\appendixtrue

@@

%\pagestyle{plain}

@@

Finite automata are precisely the class of machines that recognize regular
expressions. \cref{fig:exampleNFA} shows a non-deterministic
finite automaton (NFA) for the regular expression
$({\literal a}^{*} \otimes \literal b) \oplus \literal c$, along with a type $\Trace$, an \emph{indexed} inductive linear type of traces through this automaton. Defining an indexed inductive type can be thought
of as defining a family of mutually recursive inductive types, one for each element of the indexing type. Here $\Trace$ 
uses an index $\s : \texttt{Fin 3}$ which picks out which
state in the automaton a trace begins at --- where $\texttt{Fin 3}$ is the
finite type containing inhabitants $\{\0 , \1 , \2\}$. We can think of this as defining three mutually recursive inductive types $\Trace~\0$,
$\Trace~\1$, and $\Trace~\2$.
%
There are three kinds of constructors for $\Trace$: (1) those that
terminate traces, (2) those that correspond to transitions labeled by
a character, and (3) those that correspond to transitions labeled by
the empty string $\epsilon$. The constructor $\texttt{stop}$
terminates a trace in the accepting state $\2$. The constructors
$\texttt{1to1}$, $\texttt{1to2}$, $\texttt{0to2}$ each define a
labeled transition through the NFA, and each of these consumes a parse
of the label's character and a trace beginning at the destination of a
transition to produce a trace beginning at the source of a
transition. The constructor $\texttt{0to1}$ behaves similarly, except
its transition is labeled with the empty string $\epsilon$. Therefore,
$\texttt{0to1}$ takes in a trace beginning at state $\1$ and returns a
trace beginning at state $\0$ corresponding to the same underlying
string.
%
Lastly, we give a lambda term that constructs an accepting trace
starting at the initial state for the string \stringquote{ab}.  Later
in \Cref{sec:applications}, we will show that we can actually
construct mutually inverse functions between the regular expression
$({\literal a}^{*} \otimes \literal b) \oplus \literal c$ and its
corresponding NFA traces ($\Trace~\0$) demonstrating that the regular
expression and the automaton capture the same language. Further, since
the functions are mutually inverse, this shows they are \emph{strongly
equivalent} as grammars.

@@

The design of \theoryabbv is based on the dependent
linear-non-linear calculus (\lnld) and Lambek calculus, also known as
non-commutative linear logic.
\cite{krishnaswami_integrating_2015,lambek58}. As in \lnld,
\theoryabbv includes both non-linear dependent types, as well as
linear types, which are allowed to depend on the non-linear types, but
not on other linear types.
%
The main point of departure from \lnld's design is that, as in Lambek calculus \cite{lambek58}, the linear
typing is \emph{non-commutative} --- i.e., that exchange is not an
admissible structural rule. Furthermore, we add a general-purpose
indexed inductive linear type connective, as well as an
\emph{equalizer} type, which we will show allows us to perform
inductive proofs of equalities between linear terms.
%
Finally, while \lnld was enhanced with special connectives
inspired by separation logic to model imperative programming, we
instead add base types and axioms to the system specifically to model
formal grammars and parsing.

The formation rules for the judgments of \theoryabbv are shown in
\Cref{fig:formation}. $\Gamma$ stands for non-linear contexts; $X,Y,Z$
stand for non-linear types; $M,N$ stand for non-linear terms, these
act as in an ordinary dependent type theory; $\Delta$ stands for
linear contexts; $A,B,C$ for linear types; and, $e,f,g$ for linear
terms. These contexts, types and terms are allowed to depend on an
ambient non-linear context $\Gamma$, but note that linear types $A$
cannot depend on any \emph{linear} variables in $\Delta$. We include
definitional equality judgments for both kinds of type and term
judgments as well. Additionally, we have a judgment $\Gamma\vdash X
\isSmall$ which is used in the definition of universe types.

@@

    \inferrule{\Gamma; \Delta \vdash e : A \and \Gamma;\Delta \vdash f : A}{\Gamma;\Delta\vdash e \equiv f : A}
  \end{mathpar}
  \caption{Formation rules}
  \label{fig:formation}
\end{figure}

@@

In addition to these type-theoretic principles, we need two additional
axioms that do not generally hold in systems based on linear
logic. First, we need that additive conjunction \emph{distributes}
over additive disjunction --- e.g., in the finitary case that $0 \& A
\cong 0$ and $(A + B) \& C \cong (A \& C) + (B \& C)$. In its most
general form, the axiom says that the definable function
${\bigoplus\limits_{f : \prod_{x:X}
    Y(x)}\bigamp\limits_{x:X}A\,x\,(f\,x) \lto
  \bigamp\limits_{x:X}\bigoplus\limits_{y:Y(x)}A\,x\,y}$ has an
inverse. Second, we need that the different constructors of
$\bigoplus$ are disjoint. We add this by adding the axiom that for any
$A : X \to L$ and $x \neq x' : X$ there is a function
$\uparrow(\equalizer{b}{\sigma\,x \circ \pi_1}{\sigma\,x'\circ \pi_2}
\lto 0)$ where $b: A(x) \& A(x')$, i.e., the grammar of pairs of an
$a:A(x)$ and an $a': A(x')$ such that $\sigma\,x\,a = \sigma\,x'\,a'$
is empty.

@@

\begin{figure}
  \begin{footnotesize}
  \begin{mathpar}
    \inferrule{}{\Gamma \vdash \I \isLinTy}\and
    \inferrule{\Gamma \vdash A \isLinTy \and \Gamma \vdash B \isLinTy}{\Gamma \vdash A \otimes B \isLinTy}\and
    \inferrule{\Gamma \vdash A \isLinTy \and \Gamma \vdash B \isLinTy}{\Gamma \vdash A \lto B \isLinTy}\and
    \inferrule{\Gamma \vdash A \isLinTy \and \Gamma \vdash B \isLinTy}{\Gamma \vdash A \tol B \isLinTy}\and
    \inferrule{\Gamma,x:X \vdash A \isLinTy}{\Gamma \vdash \bigoplus\limits_{x:X} A \isLinTy}\and
    \inferrule{\Gamma,x:X \vdash A \isLinTy}{\Gamma \vdash \bigamp\limits_{x:X} A \isLinTy}\and
    \inferrule
    {\Gamma \vdash f : \ltonl {(A \lto B)} \and \Gamma \vdash g : \ltonl { (A \lto B) }}
    {\Gamma \vdash \equalizer {a}{f}{g} \isLinTy}
  \end{mathpar}
\begin{mathpar}
    \inferrule{~}{\Gamma ; a : A \vdash a : A}
    \and
    \inferrule{\Gamma ; \Delta \vdash e : B \\ \linctxwffjdg \Gamma {A \equiv B}}{\Gamma ; \Delta \vdash e : A}
    %
    \\
    %
    \inferrule{~}{\Gamma ; \cdot \vdash () : I}
    \and
    \inferrule{\Gamma ; \Delta_2 \vdash e : I \\ \Gamma ; \Delta_1,\Delta_3 \vdash e' : C}{\Gamma ; \Delta_1,\Delta_2,\Delta_3 \vdash \letin {()} e {e'} : C}
    %
    \\
    %
    \inferrule{\Gamma ; \Delta \vdash e : A \\ \Gamma ; \Delta' \vdash e' : B}{\Gamma ; \Delta, \Delta' \vdash (e , e') : A \otimes B}
    %
    \and
    %
    \inferrule{\Gamma ; \Delta_2 \vdash e : A \otimes B \\ \Gamma ; \Delta_1, a : A, b : B, \Delta_2 \vdash e' : C}{\Gamma ;  \Delta_1, \Delta_2, \Delta_3 \vdash \letin {(a , b)} e {e'} : C}
    \\
    %
    \inferrule{\Gamma ; \Delta , a : A \vdash e : B}{\Gamma ; \Delta \vdash \lamblto a e : A\lto B}
    \and
    \inferrule{\Gamma ; \Delta' \vdash e' : A \\ \Gamma ; \Delta \vdash e : A \lto B}{\Gamma ; \Delta, \Delta' \vdash \applto {e'} {e} : B}
    \\
    %
    \inferrule{\Gamma, x : X ; \Delta  \vdash e : A}
              {\Gamma ; \Delta \vdash \dlamb x e : \LinPiTy x X A}
    \and
    \inferrule{\Gamma ; \Delta \vdash e : \LinPiTy x X A \\ \Gamma \vdash M : X}{\Gamma ; \Delta \vdash e\,.\pi\,M : \subst A {M} x}
    %
    \\
    %
    \inferrule{\Gamma \vdash M : X \quad \Gamma ; \Delta \vdash e : \subst A M x}{\Gamma ; \Delta \vdash \sigma\,M\,e : \bigoplus\limits_{x:X} A}
    %
    \and
    %
    \inferrule{\Gamma ; \Delta_2 \vdash e : \bigoplus\limits_{x:X} A \quad \Gamma, x : X ; \Delta_1, a : A, \Delta_3 \vdash e' : C}{\Gamma; \Delta_1, \Delta_2, \Delta_3 \vdash \letin {\sigma\,x\,a} e {e'}: C}
    %
    \and
    %
    \inferrule
    { \Gamma ; \Delta \vdash e : A \\
      \Gamma ; \Delta \vdash \applto {f}{e} \equiv \applto {g}{e}}
    {\Gamma ; \Delta \vdash \equalizerin{e} : \equalizer {a}{f}{g}}
    %
    \and
    %
    \inferrule
    {\Gamma ; \Delta \vdash e : \equalizer{a}{f}{g}}
    {\Gamma ; \Delta \vdash \equalizerpi {e} : A}
  \end{mathpar}    
  \end{footnotesize}
  \caption{Linear types and terms (selection)}
  \label{fig:linear-types-and-terms}
\end{figure}

@@

    \inferrule*[right=Ind$\eta$]
    {\Gamma \vdash f : \prod_{x:X}\ltonl{(\el(F\,x)(A) \lto A\,x)}
     \and \Gamma \vdash e : \prod_{x:X}\ltonl{(\mu F\,x \lto A\,x)}\\\\
      \Gamma,x:X;a:\el(F\,x)(\mu F) \vdash e\,x\,(\roll a) \equiv f\,x\,(\map(F\,x)\,e) : A\,x }
    {\fold\,F\,f\equiv e' : \prod_{x:X}\ltonl{(\mu F\,x \lto A\,x)}}
  \end{mathpar}
  \end{footnotesize}
  %% \begin{footnotesize}
  %%   \begin{align*}
  %%   \mathsf{Var} &: \prod_{X:U} X \to \SPF\,X\\
  %%   \mathsf{K} &: \prod_{X:U} L \to \SPF\,X\\
  %%   \mathsf{\bigoplus} &: \prod_{X:U}\prod_{Y:U}(Y \to \SPF\,X) \to \SPF\,X\\
  %%   \mathsf{\bigamp} &: \prod_{X:U}\prod_{Y:U}(Y \to \SPF\,X) \to \SPF\,X\\
  %%   \mathsf{\otimes} &: \prod_{X:U}\SPF\,X \to \SPF\,X \to \SPF\,X\\
  %%   \el &: \prod_{X:U}\SPF\,X \to (X \to A) \to L\\
  %%   \map &: \prod_{X:U}\prod_{F : \SPF\,X}\prod_{A,B:X\to L}{\left(\prod_{x:X}\uparrow(\unquoteTy{A\,x}\lto \unquoteTy{B\,x})\right)} \to {\uparrow(\unquoteTy{\el(F)(A)} \lto \unquoteTy{\el(F)(B)})}\\\\
  %%   \roll &: \prod_{X:U}\prod_{F:X \to \SPF\,X}\prod_{x:X}\ltonl{(\el(F\,x)(\mu\,F))}\\
  %%   \fold &: \prod_{X:U}\prod_{F:X\to\SPF\,X}\prod_{A:X \to L}
  %%   \left(\prod_{x:X}\ltonl{(\el(F\,x)(A) \lto A\,x)}\right)
  %%   \to \prod_{x:X}\ltonl{(\mu F\,x \lto A\,x)}
  %% \end{align*}
  %% \end{footnotesize}
  \caption{Strictly positive functors and indexed inductive linear types}
  \label{fig:iilt}
\end{figure}

@@

%% \paragraph{Weak and Strong Equivalence}
In the theory of formal grammars, there are two different notions of
equivalence: up to weak generative capacity, meaning just which strings are
accepted by the grammar; and up to \emph{strong} generative capacity, when
the parse trees of the two grammars are isomorphic
\cite{chom1963}. Using linear types as grammars, we can define both of
these notions of equivalence in \theoryabbv.
\begin{definition}
  \label{def:weakequiv}
  Grammars $\A$ and $\B$ are \emph{weakly equivalent} if there exists
  parse transformers $\f : \ltonl{(\A \lto \B)}$ and $\g : \ltonl {(\B
    \lto \A)}$. $\A$ is a \emph{retract} of $\B$ if $\A$ and $\B$ are
  weakly equivalent and $\lambda a. g(f(a)) \equiv \lambda a.a$. They
  are \emph{strongly equivalent} if further the other composition is
  the identity, i.e., $\lambda b. f(g(b)) \equiv \lambda b.b$.
\end{definition}

@@

%% \paragraph{Ambiguity}
A formal grammar $\A$ is ambiguous if there are multiple parse trees
for the \emph{same} string $\w$.  For example, $\a \oplus \a$ is
ambiguous because there are two parses of $\stringquote{a}$,
constructed using $\inl$ and $\inr$. On the other hand, a formal
grammar is unambiguous when there is at most one parse tree for any
input string. We can capture this notion as a type in \theoryabbv in a
clever way:
\begin{definition}
  \label{def:unambig}
  A grammar $\A$ is \emph{unambiguous} if for every linear type $\B$,
  $\f : \ltonl{(\B \lto \A)}$, and $\g : \ltonl{(\B \lto \A)}$
  then $\f \equiv \g$.
\end{definition}
\cref{def:unambig} can be read more intuitively as stating that $\A$
is unambiguous if there is at most one way to transform parses of any
other grammar $\B$ into parse of $\A$. This notion of an unambiguous
type is the analog for linear types of the definition of a (homotopy)
\emph{proposition} in the terminology of homotopy type
theory\cite{hottbook}. The most basic unambiguous types are $\top$ and
$0$, and in a system of classical logic all unambiguous types would
have to be equivalent to one of these, but with our axioms we can show
also that $\I$ and literals $\literal c$ are unambiguous. To see this, first, we
establish two useful properties of unambiguity.
\begin{lemma}[\agdalogo]
  If $\B$ is unambiguous and $\A$ is a retract of $\B$ then $\A$ is unambiguous.
  If a disjunction $\bigoplus\limits_{x:X} A(x)$ is unambiguous then
  each $A(x)$ is unambiguous.
\end{lemma}
From the first principle, we can prove that $\String$ is unambiguous,
since it is a retract of $\top$. In fact, observe that if $\A$ is a
retract of $\B$ and $\B$ is unambiguous, then in fact $\A$ and $\B$
are strongly equivalent, as the equation $\lambda b. f(g(b)) \equiv
\lambda b. b$ follows because $\B$ is unambiguous. Therefore
$\String$ is also strongly equivalent to $\top$. Next, since $\String$ is
defined as a Kleene star, we can easily show that $\String \cong \I
\oplus \Char \oplus (\Char \otimes \Char \otimes \String)$. Then by
the second principle we have that $\I$ and $\Char$ and each literal
$\literal c$ are unambiguous as well.

@@

We now turn to our main task, which is using our linear type system to
implement verified parsers. Given a grammar defined as a linear type
$A$, a first attempt at defining a parser would be to implement a
function $\uparrow{(\String \lto A)}$. But since our linear functions
must be total, this means that we can construct an $A$ parse for
\emph{every} input string, which is impossible for most grammars of
interest. Instead we might try to write a partial function as a
$\ltonl{(\String \lto (A \oplus \top))}$ using the ``option''
monad. This allows for the possibility that the input string doesn't
parse, but is far too weak as a specification: we can trivially
implement a parser for any type by always returning $\inr$. The
correct notion of a parser should be one that allows for failure, but
only in the case that a parse cannot be constructed.
\begin{definition}
  A parser for a linear type $A$ is a function $\uparrow{(\String \lto
    A \oplus A_{\neg})}$, where $A_{\neg}$ is a linear type that is
  \emph{disjoint} from $A$ in that we can implement a function
  $\uparrow{(A \& A_{\neg} \lto 0)}$.
\end{definition}
Here we replace $\top$ in our partial parser type with a type
$A_{\neg}$ that we can think of as a negation of $A$. The function
$\uparrow{(A \& A_{\neg} \lto 0)}$ ensures that it is impossible for
$A$ and $A_{\neg}$ to match the same input string. This means that in
defining a parser, we will need to define a kind of negative grammar
for strings that do not parse. Fortunately, we will see that
deterministic automata naturally support such a notion with no
additional effort: the negative grammar is simply the grammar for
traces that end in a rejecting state. This follows from the following
principle, a consequence of our axiom that $\oplus$ constructors are
disjoint.
\begin{lemma}[\agdalogo]
  If $A \oplus A_{\neg}$ is unambiguous, then $A$ and $A_{\neg}$ are
  disjoint.
\end{lemma}

@@

print(*@$_D$@*) : (s : D.states) (*@$\to$@*) (*@$\uparrow$@*)(((*@$\oplus$@*)[ b : Bool ] Trace(*@$_{\D}$@*) s b) (*@$\lto$@*) String)
print(*@$_D$@*) s ((*@$\sigma$@*) b (Trace(*@$_{\D}$@*).nil .s)) = String.nil
print(*@$_D$@*) s ((*@$\sigma$@*) b (Trace(*@$_{\D}$@*).cons c (D.(*@$\delta$@*) c .s) b a trace)) =
  String.cons ((*@$\sigma$@*) c a) (print(*@$_D$@*) (D.(*@$\delta$@*) c s) ((*@$\sigma$@*) b trace))
\end{lstlisting}
\caption{Parser/printer for DFA traces}
\label{fig:printdfa}
\end{figure}

@@

To justify our assertion that \theoryabbv is a syntax for formal
grammars and parse transformers, we will now define a
\emph{denotational semantics} that makes this mathematically precise
by defining a notion of formal grammar and parse transformer then
showing that our type theory can be soundly interpreted in this
model. We then discuss how this denotational semantics provides the
basis for our prototype implementation in Agda.
